---
title: "Mini Open Methods Memo (Activity #14)"
author: "Sumaiya Azad"
date: "2025-10-20"
format:
  pdf:
    toc: false
    number-sections: true
    geometry: margin=1in
    fig-cap-location: top
    tbl-cap-location: top
    
execute:
  echo: false
  message: false
  warning: false
---

```{r setup}
# ---- Setup & packages -------------------------------------------------

library(readxl)
library(dplyr)
library(tidyr)
library(janitor)
library(knitr)
library(kableExtra)
library(dcData)
library(ggplot2)

knitr::opts_chunk$set(
  echo    = FALSE,
  message = FALSE,
  warning = FALSE
)
```

# Introduction

This memo brings together several key activities from the course into one
reproducible document. I revisit data wrangling with the Armed Forces data,
extend my baby names visualization, and return to the Box Problem from early in
the semester. I end with a brief reflection on what I feel I have learned so
far. All of the code used to generate the results in this document is collected
in a Code Appendix at the end.

# Armed Forces Data Wrangling 

In Activities #08 and #10, we used the Armed Forces data set to practice data
wrangling and create frequency tables. Here I present a revised and fully
reproducible version of that work. Instead of reading a local Excel file, I read
directly from a Google Sheets URL so that the code can run on someone else's
computer without modification.

For this memo, I focus on a narrower sub-set of soldiers:

- **Sex:** Female soldiers only  
- **Branches:** All branches represented in the table  
- **Case:** Each row in the expanded data set represents an individual soldier  

The original Excel file has multiple header rows with merged cells. I reconstruct
the header by filling in missing branch names and combining branch and sex labels
into meaningful column names. I then pivot the data into a tidy format, convert
counts to numeric values, and use `uncount()` to expand the table so that each
row corresponds to an individual soldier. Finally, I create a two-way frequency
table of **branch by rank** for female soldiers, with both counts and percentages.

```{r armed-forces-table}
# ---- Armed Forces: wrangle & frequency table (female soldiers) --------

# Download & read Excel from Google Sheets 
sheet_url <- "https://docs.google.com/spreadsheets/d/1cn4i0-ymB1ZytWXCwsJiq6fZ9PhGLUvbMBHlzqG4bwo/export?format=xlsx&gid=597536282"
temp_xlsx <- tempfile(fileext = ".xlsx")
download.file(sheet_url, destfile = temp_xlsx, mode = "wb")

# Read Excel with no column headers
df_raw <- read_excel(temp_xlsx, col_names = FALSE)

# Convert header rows into character vectors
header1 <- as.character(unlist(df_raw[2, ]))  
header2 <- as.character(unlist(df_raw[3, ]))  

# Fill branch names across merged cells
last_branch <- NA_character_
for (i in seq_along(header1)) {
  if (!is.na(header1[i]) && nzchar(header1[i])) {
    last_branch <- header1[i]
  } else {
    header1[i] <- last_branch
  }
}

# Combine into "Branch Sex"
col_names <- ifelse(
  is.na(header2) | !nzchar(header2),
  header1,
  paste(header1, header2)
)

# Remove header rows and rename first column to pay_grade
data_clean <- df_raw[-c(1:3), ]
colnames(data_clean) <- col_names
colnames(data_clean)[1] <- "pay_grade"

# Pivot into tidy format and clean up
Armed_Forces <- data_clean |>
  pivot_longer(
    cols      = -pay_grade,
    names_to  = c("branch", "sex"),
    names_sep = " ",
    values_to = "count"
  ) |>
  filter(!is.na(count)) |>
  mutate(branch = trimws(branch)) |>
  filter(sex != "Total") |>
  mutate(count = as.numeric(count)) |>
  filter(!is.na(count))

# Expand to individual soldiers and rename pay_grade to rank
Armed_Forces_Individuals <- Armed_Forces |>
  uncount(weights = count, .remove = FALSE) |>
  mutate(soldier_id = dplyr::row_number()) |>
  select(soldier_id, pay_grade, branch, sex) |>
  rename(rank = pay_grade)

# Create formatted tables by sex
make_table <- function(data, sex_label) {
  data |>
    filter(sex == sex_label) |>
    tabyl(branch, rank) |>
    adorn_totals(where = c("row", "col")) |>
    adorn_percentages("all") |>
    adorn_pct_formatting(digits = 1) |>
    adorn_ns(position = "front") |>
    kbl(
      caption = paste(sex_label, "soldiers by branch and rank"),
      align   = "c",
      booktabs = TRUE
    ) |>
    kable_styling(
      position      = "center",
      latex_options = c("hold_position", "scale_down")
    )
}

female_table <- make_table(Armed_Forces_Individuals, "Female")

female_table
```

Table 1 summarizes the distribution of **female soldiers** by branch and rank.
Each cell contains the number of soldiers and the corresponding overall
percentage. The totals row and column provide convenient summaries across all
branches and all ranks. We can see that some branches have a higher number of
female soldiers overall, and within those branches certain ranks are more common
than others. This suggests that the relationship between branch and rank is not
uniform even within the female sub-group, and the table helps identify which
branch–rank combinations are most prevalent.

# Popularity of Baby Names over the Years

The `BabyNames` data set from the `dcData` package contains counts of how many
babies received each name by year and sex in the United States. For this memo I
chose to highlight the **top three female names** and **top three male names**
based on total counts across all available years.

I first identify the top three names for each sex, then create a data set that
contains only those names. In the plot, I distinguish sexes using **line type**:
dashed lines for girls and solid lines for boys. Using both color and line type
helps make the figure more accessible.

```{r baby-names-plot, fig.cap="Figure 1. Popularity of the top three girl and boy names in the U.S. over time.", fig.alt="A time series line chart showing yearly counts of babies with the top three girl names and the top three boy names in the U.S. Each name is shown as a colored line; dashed lines represent girls and solid lines represent boys."}
# ---- Popular Baby Names: top 3 girl & boy names -----------------------

data("BabyNames")

# Top 3 female names by total count
top_female <- BabyNames |>
  filter(sex == "F") |>
  group_by(name) |>
  summarise(total = sum(count, na.rm = TRUE), .groups = "drop") |>
  arrange(desc(total)) |>
  slice_head(n = 3) |>
  pull(name)

# Top 3 male names by total count
top_male <- BabyNames |>
  filter(sex == "M") |>
  group_by(name) |>
  summarise(total = sum(count, na.rm = TRUE), .groups = "drop") |>
  arrange(desc(total)) |>
  slice_head(n = 3) |>
  pull(name)

# Data for the selected names
baby_data <- BabyNames |>
  filter(name %in% c(top_female, top_male)) |>
  mutate(line_type = ifelse(sex == "F", "dashed", "solid"))

# Create visualization
ggplot(
  data    = baby_data,
  mapping = aes(
    x        = year,
    y        = count,
    colour   = name,
    linetype = line_type
  )
) +
  geom_line(linewidth = 0.9) +
  scale_linetype_identity() +
  labs(
    title    = "Popularity of Top 3 Girl and Boy Names in the U.S. Over Time",
    subtitle = "Dashed lines represent girls; solid lines represent boys",
    x        = "Year",
    y        = "Number of Babies with Name",
    colour   = "Name",
    caption  = "Source: dcData::BabyNames dataset (U.S. Social Security Administration)"
  ) +
  theme_minimal()
```

Figure 1 shows how the popularity of the top names for girls and boys changes
over time. Some names dominate early decades and then decline, while others
become more popular only in recent years. The dashed versus solid line types
make it easy to see differences between female and male usage. I chose to use
the top three names for each sex because they provide clear trends in the data
and highlight how naming patterns evolve across the years.

# Plotting a Mathematical Function: The Box Problem

In the Box Problem from Activity #04, we start with a rectangular piece of
paper, cut out equal squares of side length \(x\) from each corner, and fold up
the resulting flaps to form an open-top box. If the original sheet has length
\(L\) and width \(W\), then the volume of the resulting box is

\[
V(x) = x \times (L - 2x) \times (W - 2x).
\]

For this memo I use a sheet that is **36 inches by 48 inches**, taking
\(L = 48\) inches and \(W = 36\) inches. The function below implements this
formula, and I use `{ggplot2}` with `stat_function()` to plot the volume as a
function of the cut size \(x\).

```{r box-plot, fig.cap="Figure 2. Volume of an open-top box made from a 36 by 48 inch sheet as a function of cut size.", fig.alt="A smooth curve showing the volume of an open-top box constructed from a 36 by 48 inch sheet as the corner cut size increases from 0 to 18 inches. The curve starts at zero, rises to a single peak at a mid-range cut size, then decreases back toward zero."}
# ---- Box Problem: function & stat_function plot ----------------------

# Volume function for a sheet with length = 48 in and width = 36 in
volume_box <- function(x, length_in = 48, width_in = 36) {
  x * (length_in - 2 * x) * (width_in - 2 * x)
}

# Domain for x: from 0 up to half of the smaller side (36 / 2 = 18)
box_domain <- data.frame(
  x = c(0, 18)
)

# Create visualization 
ggplot(
  data    = box_domain,
  mapping = aes(x = x)
) +
  stat_function(
    fun = volume_box,
    n   = 501
  ) +
  labs(
    title = "Volume of the Box vs. Cut Size",
    x     = "Cut size x (inches)",
    y     = "Volume V(x) (cubic inches)"
  ) +
  theme_minimal()
```

Figure 2 shows that the volume starts at zero when \(x = 0\), increases to a
single peak, and then decreases back toward zero as the cut size becomes too
large. This pattern reflects the trade-off between increasing the height of the
box (with larger cuts) and shrinking the base dimensions
\((L - 2x)\) and \((W - 2x)\). The maximum volume occurs at an intermediate cut
size (approximately 6–7 inches), which yields a volume of around
5,000–5,300 cubic inches. In words, cutting out squares of about 6–7 inches from
each corner of the 36 by 48 inch sheet produces the box with the largest
possible volume.

# What I Feel I’ve Learned So Far

So far in this course I feel that I have:

- **Learned how to write reproducible code** that can be run on someone else’s
  computer. Moving my Armed Forces wrangling code away from a local file path
  and reading directly from a shared Google Sheet was a concrete step toward
  reproducibility.
- **Strengthened my data wrangling skills**, especially with `pivot_longer()`,
  `filter()`, and the `{janitor}` functions such as `tabyl()` and
  `adorn_*()`. Working through the messy Armed Forces headers made me more
  comfortable with real-world, imperfect data.
- **Improved my visualization practice** by using `{ggplot2}` and thinking about
  effective graphics. In the baby names plot, I used line type in addition to
  color to distinguish sexes, which helps with accessibility and avoids relying
  on color alone.

Overall, I feel much more comfortable with R syntax and with the full data 
analysis cycle. The Plan–Code–Improve–Polish framework has helped me structure 
my workflow more intentionally. Over time I have gradually gained more 
confidence in my data wrangling and visualization skills, and I’m working 
towards becoming more familiar with producing and publishing reproducible work.

# Code Appendix

All R code used to generate the tables and figures in this document is collected
here.

```{r code-appendix, echo=TRUE, eval=FALSE}
# ---- Setup & packages -------------------------------------------------

library(readxl)
library(dplyr)
library(tidyr)
library(janitor)
library(knitr)
library(kableExtra)
library(dcData)
library(ggplot2)

knitr::opts_chunk$set(
  echo    = FALSE,
  message = FALSE,
  warning = FALSE
)

# ---- Armed Forces: wrangle & frequency table (female soldiers) --------

sheet_url <- "https://docs.google.com/spreadsheets/d/1cn4i0-ymB1ZytWXCwsJiq6fZ9P
hGLUvbMBHlzqG4bwo/export?format=xlsx&gid=597536282"
temp_xlsx <- tempfile(fileext = ".xlsx")
download.file(sheet_url, destfile = temp_xlsx, mode = "wb")

df_raw <- read_excel(temp_xlsx, col_names = FALSE)

header1 <- as.character(unlist(df_raw[2, ]))
header2 <- as.character(unlist(df_raw[3, ]))

last_branch <- NA_character_
for (i in seq_along(header1)) {
  if (!is.na(header1[i]) && nzchar(header1[i])) {
    last_branch <- header1[i]
  } else {
    header1[i] <- last_branch
  }
}

col_names <- ifelse(
  is.na(header2) | !nzchar(header2),
  header1,
  paste(header1, header2)
)

data_clean <- df_raw[-c(1:3), ]
colnames(data_clean) <- col_names
colnames(data_clean)[1] <- "pay_grade"

Armed_Forces <- data_clean |>
  pivot_longer(
    cols      = -pay_grade,
    names_to  = c("branch", "sex"),
    names_sep = " ",
    values_to = "count"
  ) |>
  filter(!is.na(count)) |>
  mutate(branch = trimws(branch)) |>
  filter(sex != "Total") |>
  mutate(count = as.numeric(count)) |>
  filter(!is.na(count))

Armed_Forces_Individuals <- Armed_Forces |>
  uncount(weights = count, .remove = FALSE) |>
  mutate(soldier_id = dplyr::row_number()) |>
  select(soldier_id, pay_grade, branch, sex) |>
  rename(rank = pay_grade)

make_table <- function(data, sex_label) {
  data |>
    filter(sex == sex_label) |>
    tabyl(branch, rank) |>
    adorn_totals(where = c("row", "col")) |>
    adorn_percentages("all") |>
    adorn_pct_formatting(digits = 1) |>
    adorn_ns(position = "front") |>
    kbl(
      caption = paste(sex_label, "soldiers by branch and rank"),
      align   = "c",
      booktabs = TRUE
    ) |>
    kable_styling(
      position      = "center",
      latex_options = c("hold_position", "scale_down")
    )
}

female_table <- make_table(Armed_Forces_Individuals, "Female")

# ---- Popular Baby Names: top 3 girl & boy names -----------------------

data("BabyNames")

top_female <- BabyNames |>
  filter(sex == "F") |>
  group_by(name) |>
  summarise(total = sum(count, na.rm = TRUE), .groups = "drop") |>
  arrange(desc(total)) |>
  slice_head(n = 3) |>
  pull(name)

top_male <- BabyNames |>
  filter(sex == "M") |>
  group_by(name) |>
  summarise(total = sum(count, na.rm = TRUE), .groups = "drop") |>
  arrange(desc(total)) |>
  slice_head(n = 3) |>
  pull(name)

baby_data <- BabyNames |>
  filter(name %in% c(top_female, top_male)) |>
  mutate(line_type = ifelse(sex == "F", "dashed", "solid"))

ggplot(
  data    = baby_data,
  mapping = aes(
    x        = year,
    y        = count,
    colour   = name,
    linetype = line_type
  )
) +
  geom_line(linewidth = 0.9) +
  scale_linetype_identity() +
  labs(
    title    = "Popularity of Top 3 Girl and Boy Names in the U.S. Over Time",
    subtitle = "Dashed lines represent girls; solid lines represent boys",
    x        = "Year",
    y        = "Number of Babies with Name",
    colour   = "Name",
    caption  = "Source: dcData::BabyNames dataset (U.S. Social Security 
    Administration)"
  ) +
  theme_minimal()

# ---- Box Problem: function & stat_function plot ----------------------

volume_box <- function(x, length_in = 48, width_in = 36) {
  x * (length_in - 2 * x) * (width_in - 2 * x)
}

box_domain <- data.frame(
  x = c(0, 18)
)

ggplot(
  data    = box_domain,
  mapping = aes(x = x)
) +
  stat_function(
    fun = volume_box,
    n   = 501
  ) +
  labs(
    title = "Volume of the Box vs. Cut Size",
    x     = "Cut size x (inches)",
    y     = "Volume V(x) (cubic inches)"
  ) +
  theme_minimal()
```
